---
title: "p8105_hw3_ns3782"
author: "NSK"
date: "2024-10-13"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem 1

```{r problem_1}

# Load libraries and import data

library(p8105.datasets)
library(tidyverse)

data("ny_noaa")


```

## Dataset Overview

The NY NOAA dataset contains `r nrow(ny_noaa)` observations of `r ncol(ny_noaa)` weather related variables. The date range of the observations is from `r min(ny_noaa$date)` to `r max(ny_noaa$date)`. 

Key variables and their types include:

* id: `r class(ny_noaa$id)` (weather station identifier)
* date: `r class(ny_noaa$date)` (YYYY-MM-DD)
* prcp: `r class(ny_noaa$prcp)` (precipitation)
* snow: `r class(ny_noaa$snow)` (snowfall)
* snwd: `r class(ny_noaa$snwd)` (snow depth)
* tmax: `r class(ny_noaa$tmax)` (maximum temperature)
* tmin: `r class(ny_noaa$tmin)` (minimum temperature)


Missing data is an issue in this dataset. The total number of missing values across all variables is `r sum(is.na(ny_noaa))`, which represents `r round(sum(is.na(ny_noaa)) / (nrow(ny_noaa) * ncol(ny_noaa)) * 100, 2)`% of all observations in the dataset. 

For the id variable, there are `r sum(is.na(ny_noaa$id))` missing values (`r round(sum(is.na(ny_noaa$id)) / nrow(ny_noaa) * 100, 2)`% of total observations). The date variable has `r sum(is.na(ny_noaa$date))` missing values (`r round(sum(is.na(ny_noaa$date)) / nrow(ny_noaa) * 100, 2)`% of total observations). Precipitation (prcp) has `r sum(is.na(ny_noaa$prcp))` missing values (`r round(sum(is.na(ny_noaa$prcp)) / nrow(ny_noaa) * 100, 2)`% of total observations). For snowfall (snow), there are `r sum(is.na(ny_noaa$snow))` missing values (`r round(sum(is.na(ny_noaa$snow)) / nrow(ny_noaa) * 100, 2)`% of total observations). Snow depth (snwd) has `r sum(is.na(ny_noaa$snwd))` missing values (`r round(sum(is.na(ny_noaa$snwd)) / nrow(ny_noaa) * 100, 2)`% of total observations). Maximum temperature (tmax) has `r sum(is.na(ny_noaa$tmax))` missing values (`r round(sum(is.na(ny_noaa$tmax)) / nrow(ny_noaa) * 100, 2)`% of total observations). Minimum temperature (tmin) has `r sum(is.na(ny_noaa$tmin))` missing values (`r round(sum(is.na(ny_noaa$tmin)) / nrow(ny_noaa) * 100, 2)`% of total observations).

